---
title: "MSc_Project"
output:
  pdf_document:
    toc: false
    toc_depth: 3
    number_sections: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(rstan)
  library(bayesTFR)
  library(ggplot2)
  library(GGally)
})
rstan_options(auto_write = TRUE)
options(mc.cores = max(1L, parallel::detectCores(logical = FALSE)))

set.seed(123)

stan_file <- "stan/changepoint.stan"
null_stan_file <- "stan/nullmodel.stan"
data_file <- "data/fertility.csv"
bayestfr_dir <- "bayesTFR_output"

prior_colours <- c(UN_inflated = "olivedrab3",
                   Reference = "lightblue2",
                   PC = "royalblue2")

round_df <- function(x, digits = 3) {
  numeric_cols <- vapply(x, is.numeric, logical(1))
  x[numeric_cols] <- lapply(x[numeric_cols], round, digits = digits)
  x
}

log_mean_exp <- function(x) {
  m <- max(x)
  m + log(mean(exp(x - m)))
}

w2_distance <- function(x, y, n = 1000) {
  probs <- seq(0.001, 0.999, length.out = n)
  sqrt(mean((quantile(x, probs) - quantile(y, probs))^2))
}

sample_truncated_normal <- function(n, mean, sd, lower, upper) {
  out <- numeric(0)
  while (length(out) < n) {
    proposal <- rnorm(n, mean, sd)
    proposal <- proposal[proposal >= lower & proposal <= upper]
    out <- c(out, proposal)
  }
  out[seq_len(n)]
}

count_divergences <- function(fit) {
  sampler <- get_sampler_params(fit, inc_warmup = FALSE)
  sum(sapply(sampler, function(chain) sum(chain[, "divergent__"])))
}

summarise_prediction <- function(pred_draws, observed) {
  lower <- quantile(pred_draws, 0.025)
  upper <- quantile(pred_draws, 0.975)

  data.frame(median_prediction = median(pred_draws),
             lower = lower,
             upper = upper,
             absolute_error = abs(median(pred_draws) - observed),
             covered_95 = as.numeric(observed >= lower && observed <= upper))
}
```

# Data and model setup

The South Korean annual TFR data were initially obtained from KOSIS using its OpenAPI service. For reproducibility, the resulting dataset is included directly in the project repository as `data/fertility.csv`, and the analysis below reads this versioned file since the personal OpenAPI code cannot be shared.

```{r}
kosis <- read.csv(data_file, stringsAsFactors = FALSE)
required_columns <- c("Year", "TFR")
kosis <- kosis[complete.cases(kosis[, required_columns]), required_columns]
kosis <- kosis[order(kosis$Year), ]

year <- as.integer(kosis$Year)
f <- as.numeric(kosis$TFR)
T <- length(f)
```

This chunk plots the observed South Korean annual TFR series and highlights the recent rebound years.

```{r, fig.width=11, fig.height=8}
rebound_years <- c(2023, 2024, 2025)
rebound_index <- year %in% rebound_years

plot(year, f,
     type = "b",
     pch = 19,
     col = "black",
     xlim = c(min(year), max(year) + 4),
     ylim = c(0, max(f) + 0.25),
     xlab = "Year",
     ylab = "TFR",
     main = "South Korean total fertility rate")
grid()

lines(year[rebound_index],
      f[rebound_index],
      col = "firebrick",
      lwd = 2)
points(year[rebound_index],
       f[rebound_index],
       pch = 19,
       col = "firebrick",
       cex = 1.3)

label_x <- c(2022.2, 2024.0, 2026.35)
label_y <- c(f[year == 2023] - 0.18,
             f[year == 2024] + 0.23,
             f[year == 2025] + 0.02)
text(label_x,
     label_y,
     labels = year[rebound_index],
     col = "firebrick",
     cex = 1.4)
text(2025.1,
     max(f[rebound_index]) + 0.36,
     labels = "Recent rebound",
     col = "firebrick",
     cex = 1.45,
     font = 2)
```

## bayesTFR posterior medians

The empirical Bayes prior is constructed from posterior medians of the Phase III hyperparameters produced by `bayesTFR`. The saved `bayesTFR_output` directory is included in this project and loaded here. 

This folder contains the fitted MCMC output using `run.tfr.mcmc()` followed by `run.tfr3.mcmc()`. The full chain using `bayesTFR` package are not rerun in this document because they are computationally expensive; instead, the saved output is used only to extract the posterior medians of \(\bar\mu\), \(\sigma_\mu\), \(\bar\rho\), and \(\sigma_\rho\).

```{r, cache=TRUE}
mcmc3 <- get.tfr3.mcmc(sim.dir = bayestfr_dir)
phase3_quantiles <- summary(mcmc3)$phase3$results$quantiles
median_names <- c("mu", "sigma.mu", "rho", "sigma.rho")

bayestfr_medians <- data.frame(parameter = c("mu_bar", "sigma_mu", "rho_bar", "sigma_rho"),
                               posterior_median = as.numeric(phase3_quantiles[median_names, "50%"]))

mu_bar <- bayestfr_medians$posterior_median[1]
sigma_mu <- bayestfr_medians$posterior_median[2]
rho_bar <- bayestfr_medians$posterior_median[3]
sigma_rho <- bayestfr_medians$posterior_median[4]
```

## Stan data for the changepoint model

The following chunk prepares the data lists required by the Stan changepoint model under each prior specification.

```{r}
make_changepoint_data <- function(year_train, f_train, prior_id) {
  T_train <- length(f_train)
  tau_years_train <- 2015:2024
  tau_years_train <- tau_years_train[tau_years_train %in% year_train]
  tau_candidates_train <- match(tau_years_train, year_train)
  valid_tau <- tau_candidates_train >= 3L &
    tau_candidates_train <= (T_train - 1L)
  tau_years_train <- tau_years_train[valid_tau]
  tau_candidates_train <- as.integer(tau_candidates_train[valid_tau])

  list(T = T_train,
       f = f_train,
       N_tau = length(tau_candidates_train),
       tau_candidates = tau_candidates_train,
       prior_type = prior_id,
       mu_bar = mu_bar,
       sigma_mu = sigma_mu,
       rho_bar = rho_bar,
       sigma_rho = sigma_rho,
       lambda_pc = -log(0.10) / sqrt(-log(1 - 0.95^2)))
}

prior_ids <- c(UN_inflated = 1L, Reference = 2L, PC = 3L)
stan_data <- lapply(prior_ids, function(id) {
  make_changepoint_data(year_train = year,
                        f_train = f,
                        prior_id = id)
})

tau_years <- year[stan_data[[1]]$tau_candidates]
lambda_pc <- stan_data[[1]]$lambda_pc
```

## Fit changepoint model

```{r, cache=TRUE, message=TRUE, include=FALSE}
stan_iter <- as.integer(Sys.getenv("STAN_ITER", "4000"))
stan_warmup <- as.integer(Sys.getenv("STAN_WARMUP", "2000"))

stan_model_compiled <- stan_model(file = stan_file)

fits <- lapply(names(stan_data), function(prior_name) {
  sampling(object = stan_model_compiled,
           data = stan_data[[prior_name]],
           chains = 4,
           iter = stan_iter,
           warmup = stan_warmup,
           seed = 123,
           refresh = 100,
           control = list(adapt_delta = 0.998, max_treedepth = 10))
})
names(fits) <- names(stan_data)

draws <- lapply(fits, rstan::extract)
model_parameters <- c("alpha", "phi", "sigma1", "mu", "rho", "sigma2")
```

Here the changepoint model is refitted again using the original empirical Bayes UN prior, only to compare its MCMC diagnostics against the final UN-inflated prior. It will not be used as a final model for posterior inference.

```{r, cache=TRUE, message=TRUE, include=FALSE}
original_un_model_code <- paste(readLines(stan_file), collapse = "\n")
original_un_model_code <- sub("mu ~ normal(mu_bar, 2 * sigma_mu);",
                              "mu ~ normal(mu_bar, sigma_mu);",
                              original_un_model_code,
                              fixed = TRUE)
original_un_model_code <- sub("rho ~ normal(rho_bar, 2 * sigma_rho);",
                              "rho ~ normal(rho_bar, sigma_rho);",
                              original_un_model_code,
                              fixed = TRUE)

original_un_model_compiled <- stan_model(model_code = original_un_model_code)

original_un_fit <- sampling(object = original_un_model_compiled,
                            data = stan_data[["UN_inflated"]],
                            chains = 4,
                            iter = stan_iter,
                            warmup = stan_warmup,
                            seed = 123,
                            refresh = 100,
                            control = list(adapt_delta = 0.998, max_treedepth = 10))
```

## Fit no-changepoint null model

```{r, cache=TRUE, message=TRUE, include=FALSE}
null_model_compiled <- stan_model(file = null_stan_file)

null_fit <- sampling(object = null_model_compiled,
                     data = list(T = T, f = f),
                     chains = 4,
                     iter = stan_iter,
                     warmup = stan_warmup,
                     seed = 123,
                     refresh = 100,
                     control = list(adapt_delta = 0.998, max_treedepth = 10))

null_draws <- rstan::extract(null_fit)
```

# 1. MCMC diagnostics

```{r}
make_diagnostic_rows <- function(fit, model, prior, parameters) {
  fit_summary <- summary(fit, pars = parameters)$summary

  data.frame(model = model,
             prior = prior,
             parameter = rownames(fit_summary),
             Rhat = fit_summary[, "Rhat"],
             ESS = fit_summary[, "n_eff"],
             divergences = count_divergences(fit),
             row.names = NULL)
}

diagnostics_table <- do.call(rbind, lapply(names(fits), function(prior_name) {
  make_diagnostic_rows(fit = fits[[prior_name]],
                       model = "Changepoint",
                       prior = prior_name,
                       parameters = model_parameters)
}))

diagnostics_table <- rbind(diagnostics_table,
                           make_diagnostic_rows(fit = null_fit,
                                                model = "No changepoint",
                                                prior = "Phase II only",
                                                parameters = c("alpha", "phi", "sigma1")))
diagnostics_full_table <- diagnostics_table
diagnostics_full_table$prior <- factor(diagnostics_full_table$prior,
                                       levels = c("UN_inflated", "Reference", "PC", "Phase II only"))
diagnostics_full_table <- diagnostics_full_table[order(diagnostics_full_table$prior,
                                                       diagnostics_full_table$parameter), ]
diagnostics_full_table$prior <- as.character(diagnostics_full_table$prior)
diagnostics_full_table$divergences <- NULL

diagnostics_summary_table <- aggregate(cbind(Rhat, ESS, divergences) ~ prior,
                                       data = diagnostics_table,
                                       FUN = function(x) {
                                         c(max = max(x, na.rm = TRUE),
                                           min = min(x, na.rm = TRUE))
                                       })

diagnostics_summary_table <- data.frame(prior = diagnostics_summary_table$prior,
                                        max_Rhat = diagnostics_summary_table$Rhat[, "max"],
                                        min_ESS = diagnostics_summary_table$ESS[, "min"],
                                        divergences = diagnostics_summary_table$divergences[, "max"],
                                        row.names = NULL)

diagnostics_summary_table$prior <- factor(diagnostics_summary_table$prior,
                                          levels = c("UN_inflated", "Reference", "PC", "Phase II only"))
diagnostics_summary_table <- diagnostics_summary_table[order(diagnostics_summary_table$prior), ]
diagnostics_summary_table$prior <- as.character(diagnostics_summary_table$prior)

print(round_df(diagnostics_summary_table), row.names = FALSE)
print(round_df(diagnostics_full_table), row.names = FALSE)
```

Below compares the original UN empirical Bayes prior with the inflated version used in the final analysis.

```{r}
original_un_summary <- summary(original_un_fit, pars = model_parameters)$summary
un_inflated_summary <- summary(fits[["UN_inflated"]],
                               pars = model_parameters)$summary

un_inflation_diagnostic_table <- data.frame(prior = c("Original UN", "UN-inflated"),
                                            max_Rhat = c(max(original_un_summary[, "Rhat"], na.rm = TRUE),
                                                         max(un_inflated_summary[, "Rhat"], na.rm = TRUE)),
                                            min_ESS = c(min(original_un_summary[, "n_eff"], na.rm = TRUE),
                                                        min(un_inflated_summary[, "n_eff"], na.rm = TRUE)),
                                            divergences = c(count_divergences(original_un_fit),
                                                            count_divergences(fits[["UN_inflated"]])),
                                            row.names = NULL)

print(round_df(un_inflation_diagnostic_table), row.names = FALSE)
```

# 2. Posterior inference on the transition year

Computing the Rao--Blackwellised posterior probabilities for tau

```{r}
tau_table <- do.call(rbind, lapply(names(fits), function(prior_name) {
  data.frame(prior = prior_name,
             tau_year = tau_years,
             probability = colMeans(draws[[prior_name]]$tau_prob))
}))
```

Plotting posterior probabilities of tau under each prior specification

```{r, fig.width=10, fig.height=4}
tau_matrix <- tapply(tau_table$probability,
                     list(tau_table$prior, tau_table$tau_year),
                     sum)

prior_order <- c("UN_inflated", "Reference", "PC")
tau_matrix <- tau_matrix[prior_order, , drop = FALSE]

prior_labels <- c(UN_inflated = "UN-inflated",
                  Reference = "Reference",
                  PC = "Penalised Complexity")

tau_year_values <- as.numeric(colnames(tau_matrix))

par(mfrow = c(1, nrow(tau_matrix)),
    mar = c(4.5, 5.6, 3, 1),
    oma = c(0, 0, 0, 0),
    cex.axis = 1.25,
    cex.lab = 1.45,
    cex.main = 1.6)

for (focal_prior in rownames(tau_matrix)) {
  plot(tau_year_values,
       tau_matrix[focal_prior, ],
       type = "n",
       ylim = c(0, max(tau_matrix)),
       xlab = "Transition year",
       ylab = expression(p(tau~"|"~f)),
       main = prior_labels[focal_prior])
  grid()

  for (prior_name in rownames(tau_matrix)) {
    if (prior_name != focal_prior) {
      lines(tau_year_values,
            tau_matrix[prior_name, ],
            type = "b",
            pch = 19,
            lwd = 1.5,
            col = adjustcolor("gray60", alpha.f = 0.45))
    }
  }

  lines(tau_year_values,
        tau_matrix[focal_prior, ],
        type = "b",
        pch = 19,
        lwd = 2.8,
        col = prior_colours[focal_prior])
}
```


# 3. Prior sensitivity analysis

## 3.1 Posterior summary table for mu and rho

```{r}
phase3_summary_table <- do.call(rbind, lapply(names(fits), function(prior_name) {
  x <- summary(fits[[prior_name]],
               pars = c("mu", "rho"),
               probs = c(0.025, 0.5, 0.975))$summary

  data.frame(prior = prior_name,
             parameter = rownames(x),
             mean = x[, "mean"],
             median = x[, "50%"],
             lower_95 = x[, "2.5%"],
             upper_95 = x[, "97.5%"],
             row.names = NULL)
}))

print(round_df(phase3_summary_table), row.names = FALSE)
```

## 3.2 Posterior pairs plot for mu and rho

```{r, fig.width=7, fig.height=7}
prior_order <- c("UN_inflated", "Reference", "PC")
prior_labels <- c(UN_inflated = "UN-inflated",
                  Reference = "Reference",
                  PC = "Penalised Complexity")

pairplot_data <- do.call(rbind, lapply(prior_order, function(prior_name) {
  data.frame(mu = draws[[prior_name]]$mu,
             rho = draws[[prior_name]]$rho,
             prior = unname(prior_labels[prior_name]),
             row.names = NULL)
}))

pairplot_data$prior <- factor(pairplot_data$prior,
                              levels = unname(prior_labels[prior_order]))

ggpairs(pairplot_data,
        columns = c("mu", "rho"),
        mapping = aes(colour = prior, fill = prior),
        columnLabels = c("mu", "rho"),
        diag = list(continuous = wrap("densityDiag", alpha = 0.35)),
        lower = list(continuous = wrap("points", size = 0.7, alpha = 0.35)),
        upper = list(continuous = wrap("cor", size = 5))) +
  theme_bw(base_size = 15) +
  theme(strip.background = element_rect(fill = "gray85", colour = "gray70"),
        strip.text = element_text(face = "bold", size = 15),
        axis.title = element_text(face = "bold", size = 16),
        axis.text = element_text(size = 13),
        panel.grid = element_line(colour = "gray88", linetype = "dotted"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 13))
```

## 3.3 Prior-posterior density plots for mu and rho

```{r}
#Construct prior draws for the density plots
n_prior_draws <- max(vapply(draws,
                            function(x) length(x$mu),
                            integer(1)))

prior_samples <- lapply(names(draws), function(prior_name) {
  if (prior_name == "UN_inflated") {
    mu_prior <- sample_truncated_normal(n = n_prior_draws,
                                        mean = mu_bar,
                                        sd = 2 * sigma_mu,
                                        lower = 0,
                                        upper = 2.1)
    rho_prior <- sample_truncated_normal(n = n_prior_draws,
                                         mean = rho_bar,
                                         sd = 2 * sigma_rho,
                                         lower = 0,
                                         upper = 1)
  } else if (prior_name == "Reference") {
    mu_prior <- runif(n_prior_draws, 0, 2.1)
    rho_prior <- 1 - sqrt(1 - runif(n_prior_draws))
  } else {
    distance <- rexp(n_prior_draws, rate = lambda_pc)
    mu_prior <- sample_truncated_normal(n = n_prior_draws,
                                        mean = mu_bar,
                                        sd = sigma_mu,
                                        lower = 0,
                                        upper = 2.1)
    rho_prior <- sqrt(1 - exp(-distance^2))
  }
  sigma2_prior <- runif(n_prior_draws, 0.001, 0.5)

  data.frame(prior = prior_name,
             mu = mu_prior,
             rho = rho_prior,
             sigma2 = sigma2_prior)
})
names(prior_samples) <- names(draws)
```

Plotting the prior and posterior densities.

```{r, fig.width=10, fig.height=12}
layout(matrix(c(1, 2, 3, 4, 5, 6, 7, 7), nrow = 4, byrow = TRUE),
       heights = c(1, 1, 1, 0.16))

par(mar = c(4.7, 6.8, 1.2, 1), cex.axis = 1.55, cex.lab = 1.85, font.lab = 2)

prior_order <- c("UN_inflated", "Reference", "PC")
prior_labels <- c(UN_inflated = "UN-inflated",
                  Reference = "Reference",
                  PC = "Penalised Complexity")
chain_colours <- c("#E69F00", "#8E7CC3", "#56B4E9", "#D55E00")
names(chain_colours) <- paste0("Chain ", seq_along(chain_colours))

for (prior_name in prior_order) {
  for (parameter_name in c("mu", "rho")) {
    x_support <- if (parameter_name == "mu") c(0, 2.1) else c(0, 1)
    x_range <- if (parameter_name == "mu") c(-0.1, 2.2) else c(-0.05, 1.05)

    posterior_by_chain <- rstan::extract(fits[[prior_name]], pars = parameter_name, permuted = FALSE, inc_warmup = FALSE)
    posterior_by_chain <- posterior_by_chain[, , parameter_name]
    n_chains <- dim(posterior_by_chain)[2]
    chain_colours_i <- chain_colours[seq_len(n_chains)]

    prior_density <- density(prior_samples[[prior_name]][[parameter_name]],
                             from = x_support[1], to = x_support[2],
                             cut = 0)

    posterior_densities <- lapply(seq_len(n_chains), function(chain_i) {
      density(posterior_by_chain[, chain_i],
              from = x_support[1], to = x_support[2],
              cut = 0)
    })

    y_range <- range(prior_density$y, unlist(lapply(posterior_densities, function(x) x$y)))
    y_range <- c(0, y_range[2] * 1.05)
    x_axis_label <- if (parameter_name == "mu") {
      expression(bold("mu"))
    } else {
      expression(bold("rho"))
    }

    plot(prior_density,
         type = "n",
         xlim = x_range,
         ylim = y_range,
         xlab = x_axis_label,
         ylab = "Density",
         main = "",
         font.lab = 2)
    if (parameter_name == "mu") {
      mtext(prior_labels[prior_name],
            side = 2,
            line = 5.1,
            font = 2,
            cex = 1.65)
    }
    grid(col = "gray88", lty = "dotted")

    for (chain_i in seq_len(n_chains)) {
      density_i <- posterior_densities[[chain_i]]
      polygon(c(density_i$x, rev(density_i$x)),
              c(density_i$y, rep(0, length(density_i$y))),
              col = adjustcolor(chain_colours_i[chain_i], alpha.f = 0.35),
              border = NA)
      lines(density_i,
            col = adjustcolor(chain_colours_i[chain_i], alpha.f = 0.9),
            lwd = 1.4)
    }

    lines(prior_density, col = "gray20", lwd = 2.2, lty = 2)
    abline(v = x_support,
           col = "gray45",
           lty = 3,
           lwd = 1.3)
  }
}

par(mar = c(0, 0, 0, 0))
plot.new()
legend("center",
       legend = c("Prior", paste("Chain", seq_len(n_chains))),
       col = c("gray20", chain_colours_i),
       lty = c(2, rep(1, n_chains)),
       lwd = c(2.2, rep(1.6, n_chains)),
       horiz = TRUE,
       bty = "o",
       bg = "white",
       box.col = "gray40",
       cex = 1.6,
       seg.len = 1.8,
       x.intersp = 0.9)
```

The following chunk reports prior and posterior mass near the upper boundary.

```{r}
rho_boundary_table <- do.call(rbind, lapply(names(draws), function(prior_name) {
  rho_post <- draws[[prior_name]]$rho
  rho_prior <- prior_samples[[prior_name]]$rho

  data.frame(prior = prior_name,
             posterior_min = min(rho_post),
             posterior_max = max(rho_post),
             prior_P_rho_gt_095 = mean(rho_prior > 0.95),
             posterior_P_rho_gt_095 = mean(rho_post > 0.95),
             prior_P_rho_gt_099 = mean(rho_prior > 0.99),
             posterior_P_rho_gt_099 = mean(rho_post > 0.99))
}))

rho_boundary_report_table <- rho_boundary_table[
  ,
  c("prior", "prior_P_rho_gt_095", "posterior_P_rho_gt_095")
]

print(round_df(rho_boundary_report_table), row.names = FALSE)
```

## 3.4 Wasserstein-2 distance between posterior distributions

```{r}
prior_pairs <- combn(names(fits), 2, simplify = FALSE)
w2_parameters <- c("mu", "rho")

wasserstein_table <- do.call(rbind, lapply(w2_parameters, function(parameter_name) {
  do.call(rbind, lapply(prior_pairs, function(pair) {
    posterior_1 <- draws[[pair[1]]][[parameter_name]]
    posterior_2 <- draws[[pair[2]]][[parameter_name]]

    data.frame(parameter = parameter_name,
               prior_1 = pair[1],
               prior_2 = pair[2],
               W2_distance = w2_distance(posterior_1, posterior_2))
  }))
}))

print(round_df(wasserstein_table), row.names = FALSE)
```

# 4. Posterior predictive checks

## 4.1 One-step-ahead predictive check after 2015

```{r}
one_step_indices <- which(year > 2015 & seq_along(year) >= 3)

one_step_rows <- list()

for (prior_name in names(draws)) {
  d <- draws[[prior_name]]

  for (t in one_step_indices) {
    phase2_mean <- f[t - 1] + d$alpha + d$phi * (f[t - 1] - f[t - 2])
    phase3_mean <- (1 - d$rho) * d$mu + d$rho * f[t - 1]

    pred_mean <- ifelse(t <= d$tau_draw, phase2_mean, phase3_mean)
    pred_sd <- ifelse(t <= d$tau_draw, d$sigma1, d$sigma2)
    pred_draws <- rnorm(length(pred_mean), pred_mean, pred_sd)

    one_step_rows[[length(one_step_rows) + 1L]] <- cbind(
      data.frame(model = "Changepoint",
                 prior = prior_name,
                 Year = year[t],
                 observed = f[t]),
      summarise_prediction(pred_draws, f[t])
    )
  }
}

for (t in one_step_indices) {
  phase2_mean <- f[t - 1] + null_draws$alpha +
    null_draws$phi * (f[t - 1] - f[t - 2])
  pred_draws <- rnorm(length(phase2_mean), phase2_mean, null_draws$sigma1)

  one_step_rows[[length(one_step_rows) + 1L]] <- cbind(
    data.frame(model = "No changepoint",
               prior = "Phase II only",
               Year = year[t],
               observed = f[t]),
    summarise_prediction(pred_draws, f[t])
  )
}

one_step_table <- do.call(rbind, one_step_rows)
```

```{r}
one_step_summary_table <- aggregate(cbind(absolute_error, lower, upper, covered_95) ~ model + prior,
                                    data = one_step_table,
                                    FUN = mean)

one_step_summary_table <- one_step_summary_table[order(one_step_summary_table$absolute_error), ]
one_step_report_table <- one_step_summary_table[, c("model", "prior", "absolute_error")]

print(round_df(one_step_report_table), row.names = FALSE)
```


# 5. Model comparison using LFO-CV

```{r, cache=TRUE, include=FALSE}
lfo_validation_years <- 2018:max(year)
lfo_iter <- as.integer(Sys.getenv("LFO_STAN_ITER", "3000"))
lfo_warmup <- as.integer(Sys.getenv("LFO_STAN_WARMUP", "1500"))
lfo_chains <- as.integer(Sys.getenv("LFO_STAN_CHAINS", "4"))

lfo_rows <- list()

for (validation_year in lfo_validation_years) {
  train_index <- year < validation_year
  year_train <- year[train_index]
  f_train <- f[train_index]
  f_test <- f[year == validation_year]
  T_train <- length(f_train)

  for (prior_name in names(prior_ids)) {
    lfo_data <- make_changepoint_data(year_train = year_train,
                                      f_train = f_train,
                                      prior_id = prior_ids[prior_name])

    lfo_fit <- sampling(object = stan_model_compiled,
                        data = lfo_data,
                        chains = lfo_chains,
                        iter = lfo_iter,
                        warmup = lfo_warmup,
                        seed = 123 + validation_year + prior_ids[prior_name],
                        refresh = 0,
                        control = list(adapt_delta = 0.998, max_treedepth = 15))

    d <- rstan::extract(lfo_fit)
    n_divergent <- count_divergences(lfo_fit)
    lfo_fit_summary <- summary(lfo_fit, pars = c("alpha", "phi", "sigma1", "mu", "rho", "sigma2"))$summary
    lfo_min_ESS <- min(lfo_fit_summary[, "n_eff"], na.rm = TRUE)

    phase2_mean <- f_train[T_train] + d$alpha + d$phi * (f_train[T_train] - f_train[T_train - 1])
    phase3_mean <- (1 - d$rho) * d$mu + d$rho * f_train[T_train]

    pred_mean <- ifelse((T_train + 1L) <= d$tau_draw, phase2_mean, phase3_mean)
    pred_sd <- ifelse((T_train + 1L) <= d$tau_draw, d$sigma1, d$sigma2)
    pred_draws <- rnorm(length(pred_mean), pred_mean, pred_sd)

    lfo_rows[[length(lfo_rows) + 1L]] <- cbind(
      data.frame(model = "Changepoint",
                 prior = prior_name,
                 validation_year = validation_year,
                 observed = f_test),
      summarise_prediction(pred_draws, f_test),
      data.frame(lfo_elpd_point = log_mean_exp(dnorm(f_test, pred_mean, pred_sd, log = TRUE)),
                 min_ESS = lfo_min_ESS,
                 divergences = n_divergent)
    )
  }

  null_lfo_fit <- sampling(object = null_model_compiled,
                           data = list(T = T_train, f = f_train),
                           chains = lfo_chains,
                           iter = lfo_iter,
                           warmup = lfo_warmup,
                           seed = 456 + validation_year,
                           refresh = 0,
                           control = list(adapt_delta = 0.998, max_treedepth = 15))

  d_null <- rstan::extract(null_lfo_fit)
  n_divergent_null <- count_divergences(null_lfo_fit)
  null_lfo_fit_summary <- summary(null_lfo_fit, pars = c("alpha", "phi", "sigma1"))$summary
  null_lfo_min_ESS <- min(null_lfo_fit_summary[, "n_eff"], na.rm = TRUE)

  null_mean <- f_train[T_train] + d_null$alpha + d_null$phi * (f_train[T_train] - f_train[T_train - 1])
  null_sd <- d_null$sigma1
  null_pred_draws <- rnorm(length(null_mean), null_mean, null_sd)

  lfo_rows[[length(lfo_rows) + 1L]] <- cbind(
    data.frame(model = "No changepoint",
               prior = "Phase II only",
               validation_year = validation_year,
               observed = f_test),
    summarise_prediction(null_pred_draws, f_test),
    data.frame(lfo_elpd_point = log_mean_exp(dnorm(f_test, null_mean, null_sd, log = TRUE)),
               min_ESS = null_lfo_min_ESS,
               divergences = n_divergent_null)
  )
}

lfo_table <- do.call(rbind, lfo_rows)
```

diagnostics;

```{r}
lfo_ess_report_table <- aggregate(min_ESS ~ model + prior,
                                  data = lfo_table,
                                  FUN = min)
lfo_ess_report_table <- lfo_ess_report_table[, c("model", "prior","min_ESS")]
lfo_ess_report_table <- lfo_ess_report_table[order(lfo_ess_report_table$model,
                                                   lfo_ess_report_table$prior), ]

print(round_df(lfo_ess_report_table), row.names = FALSE)
```

Diagnostics for the reference prior refits.

```{r}
lfo_reference_diagnostic_table <- lfo_table[lfo_table$model == "Changepoint" &
                                            lfo_table$prior == "Reference",
                                            c("validation_year", "min_ESS", "divergences")]
lfo_reference_diagnostic_table <- lfo_reference_diagnostic_table[
  order(lfo_reference_diagnostic_table$validation_year), ]

print(round_df(lfo_reference_diagnostic_table), row.names = FALSE)
```

LFO-CV comparison over the full and restricted validation windows.

```{r}
full_lfo_table <- aggregate(lfo_elpd_point ~ prior,
                            data = lfo_table,
                            FUN = sum)
names(full_lfo_table)[names(full_lfo_table) == "lfo_elpd_point"] <- "full_window_LFO_ELPD"

failed_years <- c(2018, 2019, 2021, 2022)
lfo_restricted_years <- setdiff(lfo_validation_years, failed_years)
lfo_table_restricted <- lfo_table[lfo_table$validation_year %in% lfo_restricted_years, ]

restricted_lfo_table <- aggregate(lfo_elpd_point ~ prior,
                                  data = lfo_table_restricted,
                                  FUN = sum)
names(restricted_lfo_table)[names(restricted_lfo_table) == "lfo_elpd_point"] <- "restricted_window_LFO_ELPD"

lfo_comparison_table <- merge(full_lfo_table,
                              restricted_lfo_table,
                              by = "prior")
lfo_comparison_table <- lfo_comparison_table[order(lfo_comparison_table$full_window_LFO_ELPD,
                                                   decreasing = TRUE), ]

print(round_df(lfo_comparison_table), row.names = FALSE)
```

